---
title: "00_all"
format:
html:
  embed_resources = TRUE
---

# Background

Prostate cancer is one of the most common cancer types among men, and approximately 12–13% are expected to be diagnosed at some point during their lifetime. Especially once over the age of 65 where 60% of cases are diagnosed. Methods used to diagnose include Serum hemoglobin testing and prostatic acid phosphatase (PAP) testing. Both tests have been used for many years. Serum hemoglobin tests for low and high glycosylate levels in the hemoglobin, with low levels signaling anemia or chronic disease processes, which may occur in patients with advanced cancer, while high levels can be associated with certain physiological conditions. Variations in hemoglobin therefore serve as a general indicator of systemic health status. Another used indicator is prostatic acid phosphatase (PAP), which is an enzyme produced by the prostate gland. Elevated PAP levels can indicate increased activity in prostate tissue which is often a precursor for prostate cancer or other prostate related diseases. It is hence important to observe these values to determine if a patient might be sick with cancer. Today prostate-specific-antigen (PSA) are used for a more sensitive detection of early stage prostate cancer.

# Aim

This project aims to investigate the clinical data in data set. This includes correlation of markers; hemoglobin and prostatic acid phosphatase in patients, with the goal to determine if the biomarkers are a good estimate when predicting or evaluating prostate cancer. Along with this the effect of hormone treatment will also be investigated.

# Load Data

{{< include 01_load.qmd >}}

# Tidying Data

{{< include 02_tidy.qmd >}}

# Augment Data

{{< include 03_augment.qmd >}}

# Description

{{< include 04_describe.qmd >}}

# Analysis

{{< include 05_model.qmd >}}

# Survial Analysis

{{< include 06_survival_analysis.qmd >}}

# PCA Analysis

{{< include 07_PCA.qmd >}}


```{r}
#render all notebooks
library("quarto")

qmds <- list.files(pattern = "\\.qmd$")
qmds <- qmds[qmds != "00_all.qmd"] #dont render current notebook i dont know if that would create a recursive loop
qmds <- qmds[qmds != "0X_background.qmd"]

for (qmd_document in qmds) {
  quarto_render(qmd_document,
                output_format = "html",
                execute_params = list(embed_resources = TRUE))}

#move them to res dir
html_names <- sub("\\.qmd$", ".html", qmds)

#move the files
if (!dir.exists("../results")){
  dir.create("../results", recursive = TRUE)}

# move each file
file.rename(html_names, file.path("../results", basename(html_names)))
```
```{r}
#Render the presentation as well
quarto_render("../doc/presentation.qmd")#presentation shouldnt be moved according to proj description
```

