---
title: "03_augment"
format: html
---

## Load library

```{r}
library(tidyverse)
```

## Load data

```{r}
data_dir <- "data/"
tidy_file <- "02_dat_tidy.tsv.gz"

full_df <- read_tsv("../data/02_dat_tidy.tsv.gz")
glimpse(full_df)
```

## Augment data

### Create MPA variable

MAP (Mean Arterial Pressure) is the average blood pressure in a person’s arteries during one cardiac cycle (one heartbeat).

```{r}
augment_data<- full_df |>
  mutate(MPA = (sys_blood_pressure+ 2 *dia_blood_pressure) / 3)
glimpse(augment_data)       
```

### Create survival event variable

The tidy dataset provides the is_alive column as a factor with levels such as "alive" and "dead". For survival analysis, we convert this to a binary event indicator (1 = died, 0 = alive).

```{r}
augment_data <- augment_data |>
  mutate(
    event = ifelse(is_alive == "dead", 1, 0),
    follow_up_months = as.numeric(follow_up_months))

summary(augment_data$event)
```

## Save the data to drive

```{r}
#save a compressed tsv
data_dir <- "../data/"
aug_file <- "03_dat_aug.tsv.gz"

full_df |> write_tsv(file = str_c(data_dir, aug_file))
```
