---
title: "03_augment"
format:
  html:
    self-contained: TRUE
---

## Load library

```{r}
#| message: false
library("tidyverse")
source("99_proj_func.R")
```

## Load data

Loading the prepared tidy data.

```{r}
#| message: false
data_dir <- "data/"
tidy_file <- "02_dat_tidy.tsv.gz"

full_df <- read_tsv("../data/02_dat_tidy.tsv.gz")
full_df |> slice_sample(n = 10)
```

## Filtering out extreme AP values

Its noted on the Vanderbilt repo that some entries have extremely high AP values. Unfortunantly the explanation is no longer available. For this reason we choose to remove these rows entirely:

```{r}
full_df |>
  ggplot(aes(x=serum_prostatic_acid_phosphatase)) +
  geom_histogram(binwidth = 100) +
  labs(
    title = "Distribution of AP values"
  ) +
  best_theme()
```

Based on the plot we discard all entries with AP values above 250

```{r}
cutoff <- 250
full_df <- full_df |> 
  filter(serum_prostatic_acid_phosphatase <= cutoff)
```

```{r}
full_df |> slice_sample(n = 10)
```

## Preparing data for modelling

Modelling tumor size vs other factors may pose an interesting investigation and see which factors may be important for tumor development in advanced stages. We need to first convert the columns that are not already numeric to a numeric representation. This can be done in a number of ways but to keep it similar we choose that the higher the number the worse for the potion. For example normal activity is assigned zero and low activity such as stuck in bed is assigned four.

```{r}
df_num <- full_df |>
  mutate(is_alive = case_when(is_alive == "alive" ~ 0,
                              is_alive == "dead" ~ 1),
         performance = case_when(performance == "normal activity" ~ 0,
                                 performance == "in bed < 50% daytime" ~ 1,
                                 performance == "in bed > 50% daytime" ~ 2,
                                 performance == "confined to bed" ~ 4),
         medical_history = as.numeric(medical_history),
         ekg = case_when(ekg == "normal" ~ 0,
                         ekg == "benign" ~ 1,
                         ekg == "heart block or conduction def" ~ 2,
                         ekg == "rhythmic disturb & electrolyte ch" ~ 3,
                         ekg == "old MI" ~ 4,
                         ekg == "recent MI" ~ 5,
                         ekg == "rhythmic disturb & electrolyte ch" ~ 6,
                         ekg == "heart strain" ~ 7),
         bone_metastases = as.numeric(bone_metastases),
         treatment = case_when(treatment == "placebo" ~ 0,
                               treatment == "estrogen" ~ 1)
         )
  
```

Similarly to the example available at https://r4bds.github.io/lab06.html we need a long version of the data where current variable is stored as an entry.

```{r}
df_long <- df_num |>
  select(c(patient_no,stage, age, weight, follow_up_months,
                        sys_blood_pressure, dia_blood_pressure,
                        serum_hemoglobin, serum_prostatic_acid_phosphatase,
                        combined_index, size_of_primary_tumor, is_alive, performance, medical_history,ekg, bone_metastases, treatment, dosage)) |> 
  pivot_longer(cols = c(stage, age, weight, follow_up_months,
                        sys_blood_pressure, dia_blood_pressure,
                        serum_hemoglobin, serum_prostatic_acid_phosphatase,
                        combined_index, is_alive, performance, medical_history,ekg, bone_metastases, treatment, dosage),
               names_to = "variable",
               values_to = "value")

print(df_long |> slice_sample(n = 10))
```

For analysis nesting variables is required.

```{r}
df_long_nested <- df_long |>
  group_by(variable) |>
  nest()
df_long_nested |> head(10) |> print() #quick check
```

### Save this data

This data is saved as .Rdata to ensure that types are carried over

```{r}
save(df_long_nested, df_num, file = "../data/04_data_for_modelling.Rdata")
```

## Augment data for treatment class

On top of nesting variables, a data object with attributed treatment class is made. The MAP (Mean Arterial Pressure) variable is added to this set. MAP is the average blood pressure in a person’s arteries during one cardiac cycle (one heartbeat).

```{r}
augment_data <- full_df |>
  mutate(MAP = (sys_blood_pressure+ 2 *dia_blood_pressure) / 3,
         is_alive = case_when(is_alive == "alive" ~ 0,
                              is_alive == "dead" ~ 1),
         treatment_detail = case_when(
           treatment == "placebo" ~ "Placebo",
           treatment == "estrogen" & dosage == 0.2 ~ "0.2mg Estrogen",
           treatment == "estrogen" & dosage == 1.0 ~ "1.0mg Estrogen",
           treatment == "estrogen" & dosage == 5.0 ~ "5.0mg Estrogen",
           TRUE ~ NA_character_),
         treatment_detail = factor(treatment_detail, 
                                   levels = c("Placebo", 
                                              "0.2mg Estrogen", 
                                              "1.0mg Estrogen", 
                                              "5.0mg Estrogen")),)

augment_data |> slice_sample(n = 10)
```

### Saving treatment class data

```{r}
data_dir <- "../data/"
aug_tsv_file <- "03_dat_aug.tsv.gz"
augment_data |> write_tsv(file = str_c(data_dir,aug_tsv_file))
```
