---
title: "04_describe"
format: html
---

### Load libraries

```{r}
#| message: false
library("tidyverse")
library("viridis")
library("ggridges")
```

## Data description

This dataset is from Green and Bryar who have analysed data from a randomized clinical trial of prostate cancer paitients in stage 3 and 4. In this project the same data will be analysed

### Meta data

Below is the full prostate data set after tidying.

```{r}
#| message: false
prostate_tidy <- read_tsv("../data/02_dat_tidy.tsv.gz")
prostate_tidy
```

Below is distribution between the stages and status of patients during the trial. Both in graph and table count form.

```{r}
prostate_tidy |>
  mutate(stage = factor(stage),
         is_alive = factor(is_alive)) |>
  ggplot(aes(
    x = stage,
    fill = is_alive)) +
  geom_bar(
    color = 'black',
    position = "dodge",
    alpha = 0.4) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 10, hjust = 1, size = 10, vjust = 1.5), 
        axis.text.y = element_text(size = 6)) + 
  labs(x = "Stage", y = "Count", fill = "Status")
```

```{r}
#| message: false
prostate_tidy |>
  group_by(stage, is_alive) |>
  summarize(n = n())
```

The dataset consists of 502 patients, as seen above, 96 of were in stage 3 and alive while 193 of stage 3 had died. For stage for 52 were alive and 161 patients were dead.

### Treatment groups

```{r}
prostate_tidy |>
  mutate(stage = factor(stage),
         dosage = factor(dosage)) |>
  ggplot(aes(x = stage,
         fill = dosage)) +
  geom_bar(
    color = 'black',
    position = "dodge",
    alpha = 0.4) +
  geom_hline(yintercept = 0) +
  scale_fill_viridis_d(
    option = "D",
    labels = function(x) paste0(x, " mg")
    ) +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 10, hjust = 1, size = 10, vjust = 1.5), 
        axis.text.y = element_text(size = 6)) + 
  labs(x = "Stage", y = "Count", fill = "Dosage")
```

We can see there the treatment groups are very evenly distributed.

```{r}
prostate_tidy |>
  mutate(is_alive = factor(is_alive),
         stage = factor(stage)) |>
  ggplot(aes(x = size_of_primary_tumor, y = is_alive, fill = is_alive)) +
  geom_density_ridges(alpha = 0.6, color = "black") +
  facet_wrap(~stage) +
  scale_fill_manual(
    values = c("alive" = "#FCFDBFFF", "dead" = "#000004FF")) +
  labs(title = "Tumor size stratified by status and stage", 
       x = bquote('Tumor size in' ~ cm^2), y = "Status") +
  theme_minimal(base_family = "lato") +
  theme(legend.position = "none")
```
