---
title: "05_model"
format: html
---

```{r}
#| message: false
library("tidyverse")
library("readr")
library("broom")
source("99_proj_func.R") # custom functions and theme
```

## Load data
We load the data that was prepared for modelling:
```{r}
load("../data/04_data_for_modelling.Rdata")
```


## Modelling tumor size as a function of other factors
```{r}
df_long_nested_model <- df_long_nested |>
  mutate(model_object = map(data,
                            ~lm(size_of_primary_tumor ~ value,
                                data = .x))) |> 
  mutate(tidy_model = map(model_object, ~tidy(.x,
                                              conf.int = TRUE,
                                              conf.level = 0.95))) |> 
  unnest(cols = tidy_model) |> 
  filter(term == "value") |>  #discard intercept terms
  mutate(p_adjusted = p.adjust(p.value)) |> # correct for multiple testing
  arrange(p_adjusted) #sort pvalues to reveal highest significance

df_long_nested_model |> select(-data, -model_object, -term, -std.error, -statistic) |>  print()
```

### Volcano plot

A volcano plot shows our estimate slope vs the p-value.

```{r}
df_long_nested_model |> 
  ggplot(mapping = aes(x = estimate,
                     y = -log10(p_adjusted),
                     color = factor(p_adjusted < 0.05))) +
  geom_point(alpha = 0.9) +
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 0)+
  best_theme() +
  theme(legend.position = "none") +
  ggrepel::geom_text_repel(aes(label = case_when(p_adjusted < 0.05 ~ variable,
                                                 TRUE ~ ""))) +
  labs(
    title = "Estimate vs -log10 of adjusted p-value",
    y = "-log10 of p-values adjusted for multiple testing",
    x = "estimate")
```

### Forest plot

The forest plot shows the confidence interval of the modelled slope for all variables

```{r}
forest_plot <- df_long_nested_model |> 
  filter(p_adjusted < 0.05) |> 
  ggplot(mapping = aes(x = estimate,
                       y = fct_reorder(variable, estimate),#sort by the slope 
                       xmin = conf.low,
                       xmax = conf.high)) +
  geom_errorbar(orientation = "y") +
  geom_point() +
  geom_vline(xintercept = 0) +
  theme_minimal(base_size = 12) +
  labs(
    x = "Estimates (95% CIs)",
    y = "Variable",
    title = "Variables with significant impact on tumor size"
  ) +
  best_theme()
save_and_show("forest_plot")
```

### Multivariate analysis

Now that we've seen which variables significantly relate to `size_of_primary_tumor` we can do a multivariate model.

```{r}
model_tidy <- lm(size_of_primary_tumor ~ serum_hemoglobin + medical_history + bone_metastases,
                 data = df_num) |> 
  tidy()
print(model_tidy)
```

### Distribution of p-values

```{r}
df_long_nested_model |> select(p.value) |> 
  ggplot() +
  geom_histogram(aes(x = p.value),
                 bins = 5) +
  best_theme()
```

## Logistic regression predicting survivability (bad results, shouldnt be used in analysis)

A binary logistic regression is perfomed on the `is_alive` column. This is of limited biological relevanc but may reveal an interesting insight. interesting cols: select(is_alive, age, weight, size_of_primary_tumor,serum_prostatic_acid_phosphatase, performance)

```{r}
# Turn df is_alive column into numeric variable for modelling:
df_num <- df |> mutate(is_alive = case_when(is_alive == "alive" ~ 1,
                                            is_alive == "dead" ~ 0)) |> 
  select(is_alive, size_of_primary_tumor) |> 
  drop_na()
#Pipe it into the ggplt
df_num |> 
  ggplot(aes(y = is_alive,
             x = size_of_primary_tumor)) +
  geom_point() +
  labs(
    x = "Size of primary tumor",
    y = "Alive at time of study (1: YES, 0: NO)",
    title = "Being alive vs tumor size"
  ) +
  best_theme()
```

As we can see there seems to be a threshold for tumor size vs being alive at the time of study. As you can also hear from the formulation this is not a very interesting relationship but we will continue with the analysis anyway.

### Fitting the logistic regression

```{r}
my_glm_mdl <- glm(formula = is_alive ~ size_of_primary_tumor,
                  family = binomial(link = "logit"),
                  data = df_num)
my_glm_mdl
```

```{r}
df_num |> 
  mutate(my_glm_model = pluck(my_glm_mdl, "fitted.values")) |> 
  ggplot(aes(x = size_of_primary_tumor, y = is_alive)) +
  geom_point(alpha = 0.5,
             size = 3) +
  geom_line(aes(y = my_glm_model)) +
  labs(
    x = "Size of primary tumor",
    y = "Alive at time of study (1: YES, 0: NO)",
    title = "Being alive vs tumor size"
  ) +
  best_theme()
```

## Depature from linear models and treatment correlation: AP factor and treatment effects

```{r}
df_num
```

Reasonably if we look at just treatment data, we can explore their relationship and see if they have any effects beyond linear.

```{r}
AP_TS_log_tidy_model <- lm(
  size_of_primary_tumor ~ 
    log(serum_prostatic_acid_phosphatase) + 
    serum_hemoglobin + 
    dosage + 
    bone_metastases +
    stage,
  data = df_num)

print(AP_TS_log_tidy_model |>
        tidy())
```

Disregarding non-significant values:

```{r}
AP_TS_log_tidy_model <- lm(
  size_of_primary_tumor ~ 
    log(serum_prostatic_acid_phosphatase) +
    bone_metastases,
  data = df_num)

print(AP_TS_log_tidy_model |>
        tidy())
```

### Investigating effects on PAP:

```{r}
AP_es_log_tidy_model <- lm(
  log(serum_prostatic_acid_phosphatase) ~
    serum_hemoglobin + 
    dosage +
    bone_metastases +
    stage,
  data = df_num)

print(AP_es_log_tidy_model |> tidy())
```

Again filtering for non-significant contributors.

```{r}
AP_es_log_tidy_model <- lm(
  log(serum_prostatic_acid_phosphatase) ~
    bone_metastases +
    stage,
  data = df_num)

print(AP_es_log_tidy_model |> tidy())
```

### Plotting residuals

```{r}
resid_df <- bind_rows(
  augment(AP_TS_log_tidy_model) |> mutate(model = "Tumor size ~ log(AP)"),
  augment(AP_es_log_tidy_model) |> mutate(model = "Log(AP) ~ estrogen")
)
```

```{r}
resid_df |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.8, color = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "salmon") +
  geom_smooth(method = "loess", se = FALSE, color = "pink") +
  facet_wrap(~model, scales = "free") +
  labs(
    title = "Residual fit for model comparison",
    x = "Fitted values",
    y = "Residuals"
  ) +
  best_theme()
```

We can see these models are more promising than the linear fits, we can further see if we can improve these even further.

```{r}
AP_TS_log_tidy_model <- lm(
  size_of_primary_tumor ~ 
    log(serum_prostatic_acid_phosphatase)*dosage +
    log(serum_prostatic_acid_phosphatase)*serum_hemoglobin +
    dosage*serum_hemoglobin +
    bone_metastases,
  data = df_num)

print(AP_TS_log_tidy_model |>
        tidy())
```

```{r}
AP_es_log_tidy_model <- lm(
  log(serum_prostatic_acid_phosphatase) ~
    bone_metastases +
    stage +
    I(dosage^2) +
    I(dosage^2)*serum_hemoglobin,
  data = df_num)

print(AP_es_log_tidy_model |> tidy())
```

```{r}
BM_AP_HB_model <- lm(
  bone_metastases ~
    log(serum_prostatic_acid_phosphatase)*I(dosage**2) +
    log(serum_prostatic_acid_phosphatase)*serum_hemoglobin +
    dosage*serum_hemoglobin,
  data = df_num)

print(BM_AP_HB_model |> tidy())
```
