---
title: "05_model"
format: html
---

```{r}
#| message: false
library("tidyverse")
library("readr")
```

## Load data
```{r}
#| message: false
df <- read_tsv("../data/02_dat_tidy.tsv.gz") |> tibble()
```

```{r}
df |> slice_sample(n = 30) |> print()
```
## Preparing data for modelling

## Logistic regression predicting survivability
A binary logistic regression is perfomed on the `is_alive` column. This is of limited biological relevanc but may reveal an interesting insight.
interesting cols: select(is_alive, age, weight, size_of_primary_tumor,serum_prostatic_acid_phosphatase, performance)
```{r}
# Turn df is_alive column into numeric variable for modelling:
df_num <- df |> mutate(is_alive = case_when(is_alive == "alive" ~ 1,
                                            is_alive == "dead" ~ 0)) |> 
  select(is_alive, size_of_primary_tumor) |> 
  drop_na()
#Pipe it into the ggplt
df_num |> 
  ggplot(aes(y = is_alive,
             x = size_of_primary_tumor)) +
  geom_point()
```
### Fitting the logistic regression
```{r}
my_glm_mdl <- glm(formula = is_alive ~ size_of_primary_tumor,
                  family = binomial(link = "logit"),
                  data = df_num)
my_glm_mdl
```

```{r}
df_num |> 
  mutate(my_glm_model = pluck(my_glm_mdl, "fitted.values")) |> 
  ggplot(aes(x = size_of_primary_tumor, y = is_alive)) +
  geom_point(alpha = 0.5,
             size = 3) +
  geom_line(aes(y = my_glm_model))
```

